---
title: "Introduction to fmridesign"
description: "Quick tour of the core workflow for building fMRI design matrices."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    css: albers.css
params:
  family: "lapis"
  base_size: 13
  content_width: 80
vignette: >
  %\VignetteIndexEntry{Introduction to fmridesign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.retina = 2,
  out.width = "100%",
  fig.width = 7,
  fig.asp = 0.618,
  message = FALSE,
  warning = FALSE
)
set.seed(123)
options(pillar.sigfig = 7, width = 80)
```

```{r theme-setup, echo = FALSE, results = 'asis'}
# Palette family activation
cat(sprintf('<script>document.addEventListener("DOMContentLoaded",function(){document.body.classList.add("palette-%s");});</script>', params$family))

# Content width control
cat(sprintf('<style>:root{--content:%sch}</style>', params$content_width))
```

```{r packages, message = FALSE}
suppressPackageStartupMessages({
  library(fmridesign)
  library(ggplot2)
})

# Apply albersdown theme if available
if (requireNamespace("albersdown", quietly = TRUE)) {
  theme_set(albersdown::theme_albers(params$family, base_size = params$base_size))
}
```

## Overview

The `fmridesign` package constructs design matrices for fMRI general linear model (GLM) analysis. It provides a formula-based interface for specifying event-related designs with flexible hemodynamic response functions, parametric modulators, and baseline/nuisance regressors.

The package separates task-related regressors (event models) from baseline regressors (drift, intercepts, motion), provides informative column names, and includes visualization tools for quality control.

## The fMRI Design Matrix

In fMRI analysis, the design matrix (X) is fundamental to the general linear model:

```text
Y = Xβ + ε
```

Where:
- **Y**: Observed BOLD signal time series
- **X**: Design matrix containing predictors
- **β**: Parameter estimates (regression coefficients)
- **ε**: Error term (noise)

The design matrix typically combines task-related predictors (expected responses to events) and nuisance predictors (drift, motion, physiology). Both are essential for reliable inference.

## Package Architecture

The package uses two main components:

**Event models** (`event_model()`) specify task-related regressors. You provide onset times, event types, and optional durations. Each term can use a different HRF basis function. The model supports factorial designs, parametric modulators, and trial-specific covariates.

**Baseline models** (`baseline_model()`) specify nuisance regressors. You can add polynomial or spline drift terms, block-wise intercepts, and any additional nuisance regressors (motion, physiological noise, etc.).

## Quick Start: Complete Workflow

Here's a minimal example demonstrating the complete workflow from experimental design to design matrix:

```{r complete_example, fig.alt="Predicted BOLD regressors over time for an example event model (two runs)."}
# 1. Define the temporal structure (2 runs, 150 scans each, TR = 2s)
TR <- 2
sframe <- sampling_frame(blocklens = c(150, 150), TR = TR)

# 2. Create experimental events
set.seed(123)
# Simple two-condition design with 20 events per run
conditions <- rep(c("A", "B"), each = 10, times = 2)
onsets <- c(
  sort(runif(20, 0, 150 * TR - 10)),  # Run 1
  sort(runif(20, 0, 150 * TR - 10))   # Run 2
)
blockids <- rep(1:2, each = 20)

# 3. Build the event model
emodel <- event_model(
  onset ~ hrf(Condition, basis = "spmg1"),
  data = data.frame(
    onset = onsets,
    Condition = factor(conditions),
    block = factor(blockids)
  ),
  block = ~ block,
  sampling_frame = sframe
)

# 4. Build the baseline model
bmodel <- baseline_model(
  basis = "poly",
  degree = 3,
  sframe = sframe
)

# 5. Extract design matrices
X_task <- design_matrix(emodel)
X_baseline <- design_matrix(bmodel)

# 6. Combine into full design matrix
X_full <- cbind(X_task, X_baseline)
dim(X_full)

# 7. Visualize the complete design
plot(emodel)
```

## Understanding the Components

### Sampling Frame

The `sampling_frame` object defines the temporal structure of your fMRI data:

```{r sampling_frame}
# Single run: 200 scans, TR = 2 seconds
sframe_single <- sampling_frame(blocklens = 200, TR = 2)
print(sframe_single)

# Multiple runs: 3 runs with different lengths
sframe_multi <- sampling_frame(blocklens = c(150, 200, 150), TR = 2)
print(sframe_multi)
```

### Event Specification

Events are typically specified with the formula interface:

```{r event_specification}
# Formula interface (recommended)
emodel_formula <- event_model(
  onset ~ hrf(condition) + hrf(RT, basis = "gaussian"),
  data = data.frame(
    onset = c(10, 30, 50, 70),
    condition = factor(c("easy", "hard", "easy", "hard")),
    RT = c(0.5, 0.8, 0.4, 0.9),
    block = factor(c(1, 1, 1, 1))
  ),
  block = ~ block,
  sampling_frame = sampling_frame(100, TR = 2)
)
```

## Common Use Cases

### 1. Block Design

```{r block_design, fig.alt="Block design regressors (task and rest blocks) over time."}
# Block design with 20-second blocks
block_onsets <- seq(0, 280, by = 40)
block_conditions <- rep(c("task", "rest"), length.out = length(block_onsets))
block_durations <- rep(20, length(block_onsets))

emodel_block <- event_model(
  onset ~ hrf(condition),
  data = data.frame(
    onset = block_onsets,
    condition = factor(block_conditions),
    block = factor(rep(1, length(block_onsets)))
  ),
  block = ~ block,
  durations = block_durations,
  sampling_frame = sampling_frame(150, TR = 2)
)

plot(emodel_block)
```

### 2. Rapid Event-Related Design

```{r rapid_event}
# Rapid event-related with jittered ISI
set.seed(456)
n_events <- 60
rapid_onsets <- cumsum(runif(n_events, 2, 6))  # ISI between 2-6s
rapid_conditions <- sample(c("face", "house", "object"), n_events, replace = TRUE)

emodel_rapid <- event_model(
  onset ~ hrf(stimulus),
  data = data.frame(
    onset = rapid_onsets,
    stimulus = factor(rapid_conditions),
    block = factor(rep(1, n_events))
  ),
  block = ~ block,
  sampling_frame = sampling_frame(ceiling(max(rapid_onsets)/2) + 20, TR = 2)
)

# Check design efficiency
cor(design_matrix(emodel_rapid))
```

### 3. Parametric Modulation

```{r parametric, fig.alt="Parametric modulator (RT) regressor and conditions plotted over time."}
# Event-related design with RT modulation
set.seed(789)
n_trials <- 30
pm_onsets <- sort(runif(n_trials, 0, 200))
pm_conditions <- rep(c("congruent", "incongruent"), length.out = n_trials)
pm_RT <- rnorm(n_trials, mean = ifelse(pm_conditions == "congruent", 0.5, 0.7), sd = 0.1)

emodel_parametric <- event_model(
  onset ~ hrf(condition) + hrf(RT),
  data = data.frame(
    onset = pm_onsets,
    condition = factor(pm_conditions),
    RT = scale(pm_RT)[,1],  # Center the parametric modulator
    block = factor(rep(1, n_trials))
  ),
  block = ~ block,
  sampling_frame = sampling_frame(120, TR = 2)
)

# Visualize the parametric modulator
plot(emodel_parametric, term_name = "RT")
```

## Integration with fMRI Analysis

The design matrices created by `fmridesign` can be used with various fMRI analysis approaches:

```{r integration, eval=FALSE}
# Example: Export for use with external software
X <- cbind(design_matrix(emodel), design_matrix(bmodel))

# Save as text file for SPM, FSL, or AFNI
write.table(X, "design_matrix.txt", row.names = FALSE, col.names = TRUE)

# For use with R-based analysis
# Assuming Y is your fMRI time series data
# fit <- lm(Y ~ X - 1)  # -1 removes intercept as it's in the design

# Or use with specialized fMRI packages
# library(fmri)
# results <- fmri_glm(Y, X)
```

## Best Practices

**Inspect your design matrices visually.** Check that event timing is correct, HRF shapes are reasonable, and correlations between regressors are acceptable.

**Optimize design efficiency.** Use jittered inter-stimulus intervals and balanced event types when possible.

**Match drift models to run length.** Short runs (< 5 minutes) need fewer drift terms than long runs.

**Include nuisance regressors.** Add motion parameters, physiological noise, and artifact flags to your baseline model.

## Next Steps

For more detailed information, see the following vignettes:

- `vignette("a_03_baseline_model")`: In-depth coverage of baseline and nuisance modeling
- `vignette("a_04_event_models")`: Comprehensive guide to event model specification
- Visit the [fmrihrf package documentation](https://github.com/bbuchsbaum/fmrihrf) for HRF details

## Getting Help

If you encounter issues or have questions:

1. Check the function documentation: `?event_model`, `?baseline_model`
2. Browse all available HRFs: `list_available_hrfs()`
3. Report issues: [GitHub Issues](https://github.com/bbuchsbaum/fmridesign/issues)
